---
title: "Single-Cell Transcriptomic Map of Human and Mouse Pancreas: Extension Analysis"
format: html
---

# Paper: https://pmc.ncbi.nlm.nih.gov/articles/PMC5228327/
# GEO: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE84133
# Github: https://github.com/shenorrLabTRDF/bseqsc

This document extends the analysis of the Baron et al. (2016) dataset using modern single-cell transcriptomics tools (Seurat v5, MuSiC, DoubletFinder).

## 1. Environment Setup

```{r, eval=FALSE}
# Change path as necessary
output_path = "./data"
dir.create(output_path, showWarnings = FALSE)
setwd(output_path)

# Install required packages
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
packages <- c("Seurat", "SeuratDisk", "SeuratData", "GEOquery", "tidyverse", 
              "patchwork", "DoubletFinder", "MuSiC", "MuSiC2", "BayesPrism")
# BiocManager::install(packages) # User to run this manually
```

```{r libraries}
library(Seurat)
library(GEOquery)
library(tidyverse)
library(DoubletFinder)
library(MuSiC)
library(patchwork)
```

## 2. Data Acquisition

```{r data_acquisition, eval=FALSE}
# Download Baron et al. data from GEO (GSE84133)
if(!dir.exists("./data")) dir.create("./data")
gse <- getGEO("GSE84133", destdir = "./data")

# Download bulk data for deconvolution (GSE50244)
bulk_gse <- getGEO("GSE50244", destdir = "./data")
```

## 3. Modern Single-Cell Pipeline (Seurat v5)

### A. Preprocessing & QC
```{r preprocessing}
# Load merged counts
library(data.table)
counts_dt <- fread("./data/GSE84133/GSE84133_human_islets_all_cells.csv.gz")
genes <- counts_dt$gene
counts_matrix <- as.matrix(counts_dt[, -1])
rownames(counts_matrix) <- genes
human_sc <- CreateSeuratObject(counts = counts_matrix, project = "Pancreas_Human")

# QC Metrics
human_sc[["percent.mt"]] <- PercentageFeatureSet(human_sc, pattern = "^MT-")
VlnPlot(human_sc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# Filtering (Paper thresholds: nFeature_RNA > 200)
human_sc <- subset(human_sc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

### B. Doublet Detection (DoubletFinder)
```{r doublet_detection}
human_sc <- NormalizeData(human_sc)
human_sc <- FindVariableFeatures(human_sc)
human_sc <- ScaleData(human_sc)
human_sc <- RunPCA(human_sc)
human_sc <- RunUMAP(human_sc, dims = 1:30)

# Doublet Detection
# nExp_poi <- round(0.075 * ncol(human_sc))
# human_sc <- doubletFinder_v3(human_sc, PCs = 1:30, pN = 0.25, pK = 0.09, nExp = nExp_poi, reuse.pANN = FALSE)
# DF_name <- grep("DF.class", names(human_sc@meta.data), value = TRUE)
# human_sc <- subset(human_sc, subset = get(DF_name) == "Singlet")
```

### C. Normalization & Integration (SCTransform v2)
```{r normalization}
human_sc <- SCTransform(human_sc, vars.to.regress = "percent.mt", verbose = FALSE)
```

## 4. Advanced Deconvolution (MuSiC)

```{r deconvolution}
# Load bulk data (GSE50244)
bulk_counts <- read.table(gzfile("./data/GSE50244/GSE50244_Genes_counts_TMM_NormLength_atLeastMAF5_expressed.txt.gz"), header = TRUE, row.names = 1)

# Run MuSiC deconvolution
library(MuSiC)
library(SingleCellExperiment)

# Use the processed Seurat object for reference
sce <- as.SingleCellExperiment(human_sc)

# Harmonize gene names between bulk and single-cell
common_genes <- intersect(rownames(bulk_counts), rownames(sce))
bulk_mtx <- as.matrix(bulk_counts[common_genes, ])
sce_subset <- sce[common_genes, ]

# Run MuSiC
deconv_results <- music_prop(bulk.mtx = bulk_mtx, sc.sce = sce_subset, 
                             clusters = 'SCT_snn_res.0.8', samples = 'orig.ident')

# View Proportion Estimates
head(deconv_results$Est.prop.weighted)
```

## 5. Visualizations

```{r visual_summary}
# UMAP with clusters
DimPlot(human_sc, reduction = "umap", label = TRUE)

# Marker heatmap
top10 <- all_markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
DoHeatmap(human_sc, features = top10$gene) + NoLegend()
```

## 5. Exhaustive Extension (New Insights)

Beyond replication and normalization, we aim to gain new biological insights using systems biology approaches.

The following analysis modules are detailed in the `exhaustive_extension_insights.R` script:

1.  **Transcription Factor Activity (DoRothEA)**: Inferring master regulators (e.g., PDX1, MAFA) to identify cell states beyond simple markers.
2.  **Cell-Cell Communication (CellChat)**: Mapping ligand-receptor interactions between pancreatic endocrine, ductal, and stromal cells.
3.  **Metabolic Profiling (scMetabolism)**: Quantifying metabolic pathway activities (Glycolysis, TCA cycle) per cell type.
4.  **Trajectory Inference (Monocle 3)**: Modeling functional gradients and maturation trajectories within the islet tissue.

---
*Note: This exhaustive extension aims to transform the Baron et al. (2016) transcriptomic map into a functional regulatory and social model of the pancreas.*